Contracts for unions and intersections are hard for Nickel

### 2021-05-20 paper.tex
Pass on the optimization section

The performance of modern programs heavily relies on non-trivial optimizations
performed by the compiler or the interpreter. Even more so for functional
languages, whose operational semantics is usually far removed from the actual
primitives of modern processors: naive execution can exhibit unacceptable
slowdowns. Functional programs tend to make heavy use of functions, and a
function call is a costly operation from the point of view of low-level
execution.

One crucial optimization is inlining(Figure~\ref{fig:optimizations-inlining-ex}). Inlining eliminates a function application
by directly substituting the function for its definition at compile time (or
before execution, for interpreted language). This is especially important for
small functions that are called repeatedly.

@@

Another optimization is common subexpression elimination (CSE,
Figure~\ref{fig:optimizations-cse-ex}). CSE goes into the other direction: it
factors out equal subexpressions to avoid recomputing them multiple times.

@@

Similar transformations exist beyond CSE, such as loop-invariant code motion or
some of Haskell's let-floating~\ref{}, which amount to hoist invariant
expressions out from repeated blocks in order to avoid recomputing them. In
particular, take the example of Figure~\ref{fig:optimizations-let-floating-ex}.

@@

The partial application \lstinline+g y+ is recomputed each time \lstinline+f+ is
called. This may be costly, in particular in the presence of contracts: if
\lstinline+y+ is a list for example, and \lstinline+g+ applies a list contract
on it, the penalty could be linear in the size of \lstinline+y+. A sensible
thing to do is to factor \lstinline+g y+ out of \lstinline+f+ as in
Figure~\ref{fig:optimizations-let-floating-ex}, which is something Haskell's
let-floating transformations could indeed do, given \lstinline+g+ is pure.

The soundness of these optimizations is tied the invariance of the semantics of
programs with respect to some substitutive transformations\todo{bad naming}.
Inlining requires that one can replace the application of a function by its
body: this is $\beta$-reduction, and it is hard to think of a language where it
is not a valid transformations. On the other hand, CSE-like transformations
require an expansion property of the form

@@

In the example above, the first term prints \nickel{"hi"} two times while the
second term prints it only once. However, this property does hold for
\emph{pure} terms, that are terms without side-effects. In a pure language, or a
language with effect tracking like Haskell\todo{Cite a different language, or
remove citation}, pure terms can be safely identified for applying
optimizations. Other languages rely on under-approximations.

>>>>

The performance of modern programs heavily relies on the optimizations performed
by the compiler or the interpreter. Even more so for functional languages, whose
execution model is often far removed from actual processors, causing naive
execution to exhibit unacceptable slowdowns.

One crucial optimization is
inlining(Figure~\ref{fig:optimizations-inlining-ex}). Functional programs tend
to make heavy use of functions, and a function call is a costly operation from
the point of view of low-level execution. Inlining eliminates a function
application by directly substituting the function for its definition at compile
time (or before execution, for interpreted language). This is especially
efficient for small functions that are called repeatedly.

@@

While inlining expands an expression by substituting a definition for its value,
an opposite transformation is sometimes beneficial, for example when a composite
expression is repeated several times. This causes the same expression to be
wastefully recomputed at each occurrence. Common subexpression elimination (CSE)
consist in introducing a new identifier for this expression and using the
identifier in place of the original occurrences
(Figure~\ref{fig:optimizations-cse-ex}), computing the result once and for all.

@@

Beyond CSE, optimizations such as loop-invariant code motion or
let-floating~\ref{} apply the same principle of extracting an invariant
expression to avoid recomputation respectively across loop iterations and
function calls. For example, take the code of
Figure~\ref{fig:optimizations-let-floating-ex}.

@@

The partial application \lstinline+g y+ is recomputed each time \lstinline+f+ is
called. This may be costly, in particular in the presence of contracts: if
\lstinline+y+ is a list for example, and \lstinline+g+ applies a list contract
on it, the cost could be linear in the size of \lstinline+y+. A sensible thing
to do is to factor \lstinline+g y+ out of \lstinline+f+ as in
Figure~\ref{fig:optimizations-let-floating-ex}, which is something a
let-floating transformation could indeed do (given \lstinline+g+ is pure, as
explained below).

The soundness of all these optimizations is tied to the invariance of the
semantics of programs with respect to specific substitutions. Inlining requires
that one can replace the application of a function by its body, which is
basically $\beta$-reduction: as long as the arguments are evaluated following
the language's strategy, it is a valid transformation. On the other hand,
CSE-like transformations require an expansion property of the form

@@

In the example above, the first term prints \nickel{"hi"} two times while the
second term prints it only once. However, this property does hold for
\emph{pure} terms, that are terms without side-effects. In a pure language, or a
language with effect tracking like Haskell\todo{Cite a different language, or
remove citation}, pure terms can be safely identified for applying
optimizations.

Strikingly, we will see in section~\ref{} that the introduction of unions and
intersections can make this equivalence \todo{name it} unsound \emph{even in a
pure setting}, making the optimization of
Figure~\ref{fig:optimizations-let-floating-ex} unsound in general.
